### 2018-02-07 submission/manuscript.Rmd
change to OR

#Load in relative risk alpha diversity data tables
alpha_RR_crc_stool <- as.data.frame(read_csv("../data/process/tables/alpha_RR_composite.csv"))
alpha_RR_adn_stool <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_composite.csv"))

alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_unmatched_tissue_composite.csv"))
alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_matched_tissue_composite.csv"))

alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_tissue_composite.csv"))

#Load in relative risk select genera data tables
select_RR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_RR_stool_composite.csv"))
select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_RR_tissue_composite.csv"))

adn_select_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_stool_composite.csv"))
adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv"))

@@

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.

@@

***Lower Bacterial Diversity is Associated with Increased RR of Tumors:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.

When observing RRs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11]. Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their RR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of tumors. Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma. Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC. Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma. No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease. Overall, associations between the microbiota and carcinoma stage of disease were much stronger than with the adenoma stage. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.

@@

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected. The OTU based Random Forest Models using selected OTUs utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs. OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased RR based on P-value.

>>>>

#Load in odds ratio alpha diversity data tables
alpha_RR_crc_stool <- as.data.frame(read_csv("../data/process/tables/alpha_OR_composite.csv"))
alpha_RR_adn_stool <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_composite.csv"))

alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_unmatched_tissue_composite.csv"))
alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_matched_tissue_composite.csv"))

alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv"))

#Load in odds ratio select genera data tables
select_RR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_stool_composite.csv"))
select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_tissue_composite.csv"))

adn_select_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_OR_stool_composite.csv"))
adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_OR_tissue_composite.csv"))

@@

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in the odds ratio for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the OR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.

@@

***Lower Bacterial Diversity is Associated with Increased OR of Tumors:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in OR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.

When observing ORs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high OR for both colon tissue and stool samples [Table S10 & S11]. Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their OR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the OR of tumors. Even though mouth-associated genera increased the OR of having a carcinoma, they did not consistently increase the RR of having an adenoma. Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC. Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest OR of having a carcinoma. No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease. Overall, associations between the microbiota and carcinoma stage of disease were much stronger than with the adenoma stage. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.

@@

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected. The OTU based Random Forest Models using selected OTUs utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs. OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased OR based on P-value.
